---
title: "CRISPR analysis with Bioconductor"
author: "Jieun Jeong"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = TRUE)
```
This vignette is based on a web page of Bioinformatics Resource Centre at
Rockefeller University and it show how to use several Bioconductor
packages in the analysis of data from a CrispR screen.

## First steps

We start from setting the working directory and downloading the data if not
downloaded already.
```{r Download, message=FALSE, warning=FALSE, results='hide'}
wd <- "/Users/jieun/Work/Crispr"
wdf <- function(...) file.path(wd,...)
Download <- function(url, filename) {
  file_path <- wdf(filename)
  if (file.exists(file_path))
    return()
  options(timeout = 1000)
  download.file(url, file_path)
  unzip(file_path,exdir = wd)
}
Download("https://github.com/LewisLabUCSD/PinAPL-Py/archive/master.zip",
         "pinaplData.zip")
Download("http://pinapl-py.ucsd.edu/example-data","TestData.zip")
dN <- wdf("New_Example_1580342908_4/Analysis")
Download("http://pinapl-py.ucsd.edu/run/download/example-run","Results.zip")
```

## Quality control of fastq data

We may inspect several metrics of data quality.  For larger data sets,
we use random sampling that yields sufficiently accurate results, while
time consuming steps may be avoided if data has clearly low quality.

```{r QC with ShortRead}
require(ShortRead)
CRfile <- wdf("PinAPL-Py-master/Data/Tox-A_R01_98_S2_L008_R1_001_x.fastq.gz")
fqSample <- FastqSampler(CRfile,n=10^5)
fastq <- yield(fqSample)
fastq # check sample statistic
```
Now we use so-called accessor functions.  The choice of bin width corresponds
to the average bin quality rounded down to integer.  
```{r Accessor_functions}
readQuality <- quality(fastq)
readQualities <- alphabetScore(readQuality)
readQualities[1:14]
require(ggplot2)
toPlot <- data.frame(ReadQ=readQualities)
ggplot(toPlot,aes(x=ReadQ))+geom_histogram(binwidth=50)+theme_minimal()
```
Q40 quality can be optimistic, so other measures can be more reliable, like
how close the distribution of nucleotides on read position is close to what
we can expect.
```{r frequencies}
readSequences <- sread(fastq)
readSequences_AlpFreq <- alphabetFrequency(readSequences)
readSequences_AlpFreq[1:3,]
readSequences_AlpbyCycle <- alphabetByCycle(readSequences)
readSequences_AlpbyCycle[1:4,1:10]
```
We can use barplot to visualize the distribution of the read frequencies.
We see the reads have a constant prefix used to precipitate selectively DNA
with guiding sequences, and a variable part, guiding sequences themselves, and
effects of contamination as minorities in the prefix.
```{r freq_plot}
counts <- readSequences_AlpbyCycle[1:4,1:50]
barplot(counts, col = c("blue", "red", "green", "yellow"), 
        legend.text = rownames(counts),
        args.legend = list( x = "topright", inset = c(-0.068, 0)),
        main = "Nucleotide Counts by Cycle",
        names.arg = 1:50)
```
The initial cycle may be affected by low quality rather than contamination,
that can be checked as follows
```{r qual_per_cycle}
qualAsMatrix <- as(readQuality,"matrix")
toPlot <- colMeans(qualAsMatrix)
plot(toPlot, xlab = "cycle", ylab = "Q40")
```
The conclusion here is that quality is fine, but in mapping tolerating one
mismatch and trimming first two cycles can be considered.

## Aligning single data file

The further analysis requires to align reads to the sgRNA require
to quantify sgRNA abundance in our samples.  We start by retrieving and
indexing the reference.
```{r sgRNA_reference_GeCKO}
GeCKO <- read.delim(wdf("PinAPL-py_demo_data/GeCKOv21_Human.tsv"))
GeCKO[1:2,]
require(Biostrings)
sgRNAs <- DNAStringSet(GeCKO$seq)
names(sgRNAs) <- GeCKO$UID
writeXStringSet(sgRNAs, file=wdf("GeCKO.fa")) # sgRNA reference
Index <- wdf("GeCKO")
Index.fa <- paste0(Index,".fa")
require(Rsubread)
buildindex(Index,Index.fa, indexSplit=FALSE)
```
Now the test alignment
```{r test_alignment}
myFQs <- wdf("PinAPL-py_demo_data/Control_R1_S14_L008_R1_001_x.fastq.gz")
myMapped <- align(Index,myFQs,output_file=sub(".fastq.gz",".bam",myFQs),
                  nthreads=4,
                  unique=TRUE,
                  nBestLocations=1,
                  type = "DNA",
                  maxMismatches = 0,
                  TH1=1)
```
We got 82.7% of uniquely mapped reads, we can test what happens if we 
 allowed number a mismatch by setting `maxMismatches = 1`.  It increases
the number of mapped reads by 8000, i.e. 1.6%.  However, the reference has
sequences of length 20, so the constant prefix is soft-clipped, and one
mismatch among 20 or 19 may create data noise that could significantly
affect some genes associated with the guiding sequences in the reference.

## Quantifying alignment result for a collection of data files

To process bam files we will use packages that work with bam format without
decompressing it into sam format.  The scheme is that we create a vector of 
fasta-q files for samples in a CRISPR screen project, and then this vector
together with the reference of gRNAs used in the project we create a
data frame of counts.  For reasons we will see later we do it for two examples.

```{r FilesOfPinapl}
# Files of PinAPL demo
d <- wdf("PinAPL-Py-master/Data/*fastq.gz")
Files <- Sys.glob(d)
```

Then we apply a "data frame maker" that can be used in other projects too.

```{r Code_of_maker_of_count_matrix}
require(GenomicAlignments)
Files_2_DF <- function(Files,indexFile) {
  f_no <- length(Files)
  for (i in seq(1,f_no)) {
    fi <- Files[i]
    # co <- CrisprCounts(fi,Index)
    output <- sub("\\.f[^.]*\\.gz$",".bam", Files[i])
    align(indexFile, Files[i], output_file=output, nthreads=4, unique=TRUE,
        nBestLocations=1, type = "DNA", maxMismatches = 0, TH1=1)
    # We decided to be most conservative and require that all 20 nucleotides of
    # gRNAs are in the alignment, but one can use different criteria too.
    temp <- readGAlignments(output)
    # temp <- temp[width(temp) == 20]
    co <- data.frame(table(seqnames(temp)),row.names = "Var1")
    if (i == 1)
      DF <- co
    else
      DF <- cbind(DF,co)
  }
  return(DF)
}
```
Now we use that code
```{r Use_of_maker_of_count_matrix_1, results='hide'}
Counts <- Files_2_DF(Files,Index)
```
```{r}
head(Counts)
```

Now we can apply DESeq to find differential guides and genes.
```{r DESeq_1}
require(DESeq2)
useCol <- DataFrame(Group=factor(c("C","C","TA","TA","TB","TB"),
                    levels = c("C","TA","TB")),row.names = colnames(Counts))
dds <- DESeqDataSetFromMatrix(Counts,colData = useCol,design = ~Group)
dds <- DESeq(dds)
TAvsC <- results(dds,contrast=c("Group","TA","C"))
TAvsC <- TAvsC[order(TAvsC$pvalue),]
head(TAvsC)
```
Note that fold changes of top gRNA's are "decent" but adjusted p-values are not.
Later we will see the reason.  In general, DESeq is not recommended for CRISPR
screens because interpretation of the read counts is different, we want to know
essential genes in the screen context, not "differentially expressed".  Thus we
will try to apply recommended
```{r CRISPRcleanR_1}
require(CRISPRcleanR)
data(GeCKO_Library_v2) # GeCKO reference processed for use with CRISPRcleanR
```
With CRISPRcleanR (ccr from now) annotation for our CRISPR platform, we edit 
Counts to Counts.ccr, a format required by ccr.  Some sgRNA rows will be
removed, 1000 of them are for control sgRNAs, and a few for genes.  The issue
of annotating sgRNAs for those genes will be left for later and now we will
remove them from consideration:
```{r making_Counts.ccr}
Counts.ccr <- Counts[rownames(GeCKO_Library_v2),]
Counts.ccr <- cbind(GeCKO_Library_v2[,c("CODE","GENES")],Counts.ccr)
#normANDfcs<-ccr.NormfoldChanges(Dframe=Counts.ccr, min_reads=10,
#                                display=TRUE, saveToFig=FALSE,
#                                libraryAnnotation = GeCKO_Library_v2,
#                                ncontrols = 2)
```
It does not work.  Probably the issue lies in very poor correlation of control
samples, i.e. 0.13, compared to correlations within other pairs of replicates,
0.96 and 0.999.
```{r Corellation}
cor(Counts.ccr[,3:8])
```
To study subsequent steps in the analysis, we consider another data set
```{r KY_data}
data(KY_Library_v1.0) # sequence ids, annotations, sequence
d <- file.path(system.file("extdata",package = "CRISPRcleanR"),"*fq.gz")
# copied these files to my directory
d <- wdf("KY_extdata/*gz")
Files <- Sys.glob(d)
basename(Files)
d <- file.path(system.file("data",package = "CRISPRcleanR"))
data("KY_Library_v1.0.RData")
# create fasta of sequences in KY_Library_v1.0 and build the index
Index <- wdf("KY_Index")
Index.fa <- wdf("KY_Index.fa")
tempDF <- data.frame(ID = rownames(KY_Library_v1.0), seq = KY_Library_v1.0$seq)
faEntries <- paste0(">", tempDF$ID, "\n", tempDF$seq)
writeLines(faEntries,wdf("KY_Index.fa"))
buildindex(Index,Index.fa, indexSplit=FALSE)
```
We repeat the construction of a data frame of read counts
```{r KY_Index, results='hide'}
Counts <- Files_2_DF(Files,Index)
colnames(Counts) <- c("Control","Sample1","Sample2")
Counts$gene <- KY_Library_v1.0$GENES
Counts <- Counts[,c(length(Files)+1,1:length(Files))]
# change of format required by CRISPRclearR
Counts <- tibble::rownames_to_column(Counts, var="sgRNA")
write.table(Counts,wdf("KY_Counts"), sep='\t', quote=FALSE, row.names=F) 
```
Testing Files_2_DF and the second data set was very helpful, because the code
was corrected: fastaq file have two possible suffices: .fq and .fastaq and
converting the name to the name of .bam file had to be altered.

## Finding essentiality (loss of fitness) and gain of fitness

The goal of CRISPR screen is to identify two types of interesting genes.
The loss of fitness genes (in extreme case, essential) have smaller counts
in knockout sample compared with control, and gain of fitness genes, have
larger count.  In tumors, the latter may be tumor supresing genes, typically,
few but of special interest in designing therapies.

Without normalization, fold changes may have show artifacts caused by
several mechanism, for that reason packages like CRISPRcleanR can be used
to remove them, e.g. using the following function

```{r Pipeline_test, error=TRUE}
ccr.AnalysisPipeline(
  file_counts = wdf("KY_Counts.tsv"),
  outdir= wdf("KY_pipeline/"),
  EXPname = "KY",
  library_builtin = "KY_Library_v1.0",
  run_mageck = FALSE,
  ncontrols = 1
)
```
This does not work either, this data set fails the requirement that 80% of
the library sgRNA has at least 30 reads in control.  We can check what threshold
lower than 30 satisfies 80% criteria.
```{r 80percent_check}
pr <- function(...) print(paste(...))
r <- dim(Counts)[1]*0.8
pr('required at least',r,'in every column, trying threshold 6')
colSums(Counts[,3:5] >= 6)
pr('trying threshold 5')
colSums(Counts[,3:5] >= 5)
pr('threshold 5 almost satisfies 80%, but is 6 times to small')
pr('sums of readcounts 6 times too small, and they are')
colSums(Counts[3:5])
```

Actually, the fastaq files had 1 million fragments each, not all of them
were mapped, so we need 6 million fragments or more in fastaq files.  Let's 
test an alternative example mapped to the same set of sgRNAs
```{r HT29_data, error=TRUE}
fn <- file.path( system.file("extdata", package = "CRISPRcleanR"), 
                 "HT-29_counts.tsv")
Counts <- read.delim(fn)
dim(Counts)
colSums(Counts[3:6])/1000/1000
```
We see that sums of read counts are ca. 50 times larger than the minimum we
estimated (we showed sums in millions).  We want at most 18,000 entries below
30 in each column
```{r entries_under_30}
colSums(Counts[,3:6] < 30)
```
As expected, the critical statistics are small enough.
In the second workout we will see how to use pipeline results.
